---
title: "Data loader"
author: "NSH"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r, message=FALSE, include=FALSE}
if (!require(readxl)) install.packages("readxl")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(BiocManager)) install.packages("BiocManager")
if (!require(SummarizedExperiment)) BiocManager::install("SummarizedExperiment")
if (!require(biomaRt)) BiocManager::install("biomaRt")

source("scripts/utils/utils.R")
```

# Load data

The below chunk loads the pipeline outputs from nf-core as SummarizedExperiment.
Additionally, ensembl databases for human and mouse are retrieved.

```{r}
salmon_se_hs <- readRDS("data/raw_data/salmon/Hs_inflamm_salmon.merged.gene_counts_length_scaled.rds")
salmon_se_mm <- readRDS("data/raw_data/salmon/Mm_inflamm_salmon.merged.gene_counts_length_scaled.rds")

ensembl_human <- useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl", version = 107)
ensembl_mouse <- useEnsembl(biomart="ensembl", dataset="mmusculus_gene_ensembl", version = 107)
# rownames(arch_hs_se) <- rowData(arch_hs_se)$gene_name
```

Here, metadata is loaded, QC failing samples are removed and the SummarizedExperiment objects are updated accordingly.

```{r}
# load data
coldata_cross_updated <- data.frame(
  read_xlsx("./data/metadata/arch_annotation.xlsx",na = "NA", sheet = "inflammatory_comps"))

# take care of identical entries for with multiple runs per experiment
coldata_cross_updated <- coldata_cross_updated[!duplicated(coldata_cross_updated$SRX_accession),]
rownames(coldata_cross_updated) <- coldata_cross_updated$SRX_accession

coldata_cross_updated[["inflammatory_analysis"]] <- as.logical(coldata_cross_updated[["inflammatory_analysis"]])

# split by species
coldata_hs <- coldata_cross_updated[coldata_cross_updated[["species"]] == "Hs",]
coldata_mm <- coldata_cross_updated[coldata_cross_updated[["species"]] == "Mm",]

# match the order
coldata_hs <- coldata_hs[colnames(salmon_se_hs), ]
coldata_mm <- coldata_mm[colnames(salmon_se_mm), ]

# check if colnames of assay match rownames of colData
if (any(!rownames(coldata_hs) == colnames(assay(salmon_se_hs))) |
    any(!rownames(coldata_mm) == colnames(assay(salmon_se_mm)))) {
      warning ("ColData mismatch. Result most likely corrupted")
}

# remove QC failing samples
failed_samples <- c("SRX5391730",
                    "SRX5391731",
                    "SRX5391732",
                    "SRX5391733",
                    "SRX5391734",
                    "SRX5391735")

all_hs_samples <- rownames(coldata_hs)
ok_samples <- setdiff(all_hs_samples, failed_samples)

coldata_hs <- coldata_hs[ok_samples, ]
salmon_se_hs <- salmon_se_hs[, ok_samples]

stopifnot(all(rownames(coldata_hs) == colnames(salmon_se_hs)))
stopifnot(all(!failed_samples %in% rownames(coldata_hs)))
stopifnot(all(ok_samples %in% rownames(coldata_hs)))
```


# Gene mapping and filtering

The below code performs a mapping step to update each gene's primary identifier 
to symbols. Additionally genes are subset to only include protein coding genes.

```{r}
###############################
# ENSEMBL MAPPING + FILTERING #
###############################

#########
# HUMAN #
#########

# RETRIEVE ENSEMBL DATA
# ensembl dbs have been retrieved above

# define the information to query from biomart
attributes_human_query <- c(
  "ensembl_gene_id",
  "external_gene_name"
)

# retrieve biomart information on protein coding genes
protein_coding_genes_human <- getBM(attributes = attributes_human_query,
                              filters = c("biotype"),
                              values = list("protein_coding"),
                              mart = ensembl_human)
# COUNTS
# filter ENSEMBL IDs for protein coding genes
ensembl_counts_human_pc <- assay(salmon_se_hs, "counts") %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  mutate(ensembl_gene_id = str_remove(ensembl_gene_id, "\\..*$")) %>%
  filter(ensembl_gene_id %in% protein_coding_genes_human$ensembl_gene_id) %>%
  column_to_rownames("ensembl_gene_id")

# filter and map to genenames
gene_counts_human_pc <- assay(salmon_se_hs, "counts") %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  mutate(ensembl_gene_id = str_remove(ensembl_gene_id, "\\..*$")) %>%
  # join (inner join includes filtering) and remove empty symbols
  inner_join(protein_coding_genes_human %>%
               filter(external_gene_name != ""),
             by = "ensembl_gene_id") %>%
  # handle ambiguously mapped genes
  group_by(ensembl_gene_id) %>%
  mutate(n_gene_per_ensembl = length(external_gene_name)) %>%
  group_by(external_gene_name) %>%
  mutate(n_ensembl_per_gene = length(ensembl_gene_id)) %>%
  ungroup() %>%
  # remove genes with ensembl id mapping to multiple or no genes
  filter(n_gene_per_ensembl == 1) %>%
  # take the sum of symbols that have multiple corresponding ensembl genes
  pivot_longer(cols = -c("ensembl_gene_id",
                         "external_gene_name",
                         "n_gene_per_ensembl",
                         "n_ensembl_per_gene"),
               names_to = "sample",
               values_to = "counts") %>%
  group_by(external_gene_name, sample) %>%
  mutate(counts_sum = sum(counts)) %>%
  dplyr::select(external_gene_name, sample, counts_sum) %>%
  distinct() %>%
  pivot_wider(id_cols = c("external_gene_name"),
              values_from = "counts_sum",
              names_from = "sample") %>%
  arrange(external_gene_name) %>%
  column_to_rownames("external_gene_name")


# ABUNDANCE
# filter ENSEMBL IDs for protein coding genes
ensembl_abundance_human_pc <- assay(salmon_se_hs, "abundance") %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  mutate(ensembl_gene_id = str_remove(ensembl_gene_id, "\\..*$")) %>%
  filter(ensembl_gene_id %in% protein_coding_genes_human$ensembl_gene_id) %>%
  column_to_rownames("ensembl_gene_id")

# filter and map to genenames
gene_abundance_human_pc <- assay(salmon_se_hs, "abundance") %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  mutate(ensembl_gene_id = str_remove(ensembl_gene_id, "\\..*$")) %>%
  # join (inner join includes filtering) and remove empty symbols
  inner_join(protein_coding_genes_human %>%
               filter(external_gene_name != ""),
             by = "ensembl_gene_id") %>%
  # handle ambiguously mapped genes
  group_by(ensembl_gene_id) %>%
  mutate(n_gene_per_ensembl = length(external_gene_name)) %>%
  group_by(external_gene_name) %>%
  mutate(n_ensembl_per_gene = length(ensembl_gene_id)) %>%
  ungroup() %>%
  # remove genes with ensembl id mapping to multiple or no genes
  filter(n_gene_per_ensembl == 1) %>%
  # take the sum of symbols that have multiple corresponding ensembl genes
  pivot_longer(cols = -c("ensembl_gene_id",
                         "external_gene_name",
                         "n_gene_per_ensembl",
                         "n_ensembl_per_gene"),
               names_to = "sample",
               values_to = "abundance") %>%
  group_by(external_gene_name, sample) %>%
  mutate(abundance_sum = sum(abundance)) %>%
  dplyr::select(external_gene_name, sample, abundance_sum) %>%
  distinct() %>%
  pivot_wider(id_cols = c("external_gene_name"),
              values_from = "abundance_sum",
              names_from = "sample") %>%
  arrange(external_gene_name) %>%
  column_to_rownames("external_gene_name")


rowdata_hs <- protein_coding_genes_human %>%
  filter(external_gene_name != "",
         external_gene_name %in% rownames(gene_counts_human_pc),
         ensembl_gene_id %in% str_remove(rownames(ensembl_counts_human_pc), "\\..*")) %>%
  group_by(external_gene_name) %>%
  summarise(mapped_ensembl_ids = paste(ensembl_gene_id, collapse = ", "),
            n_mapped = n()) %>%
  ungroup() %>%
  arrange(external_gene_name) %>%
  DataFrame()

stopifnot(all(rownames(gene_counts_human_pc) == rowdata_hs$external_gene_name))
stopifnot(all(rownames(gene_abundance_human_pc) == rowdata_hs$external_gene_name))

stopifnot(all(colnames(gene_counts_human_pc) == rownames(coldata_hs)))
stopifnot(all(colnames(gene_abundance_human_pc) == rownames(coldata_hs)))

salmon_se_hs_symbol <- SummarizedExperiment(
  assays = list(counts = gene_counts_human_pc,
                abundance = gene_abundance_human_pc),
  colData = coldata_hs,
  rowData = rowdata_hs
)


#########
# MOUSE #
#########

# RETRIEVE ENSEMBL DATA
# ensembl dbs have been retrieved above

# define the information to query from biomart
attributes_mouse_query <- c(
  "ensembl_gene_id",
  "external_gene_name"
)

# retrieve biomart information on protein coding genes
protein_coding_genes_mouse <- getBM(attributes = attributes_mouse_query,
                              filters = c("biotype"),
                              values = list("protein_coding"),
                              mart = ensembl_mouse)

# COUNTS
# filter ENSEMBL IDs for protein coding genes
ensembl_counts_mouse_pc <- assay(salmon_se_mm, "counts") %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  mutate(ensembl_gene_id = str_remove(ensembl_gene_id, "\\..*$")) %>%
  filter(ensembl_gene_id %in% protein_coding_genes_mouse$ensembl_gene_id) %>%
  column_to_rownames("ensembl_gene_id")

# filter and map to genenames
gene_counts_mouse_pc <- assay(salmon_se_mm, "counts") %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  mutate(ensembl_gene_id = str_remove(ensembl_gene_id, "\\..*$")) %>%
  # join (inner join includes filtering) and remove empty symbols
  inner_join(protein_coding_genes_mouse %>%
               filter(external_gene_name != ""),
             by = "ensembl_gene_id") %>%
  # handle ambiguously mapped genes
  group_by(ensembl_gene_id) %>%
  mutate(n_gene_per_ensembl = length(external_gene_name)) %>%
  group_by(external_gene_name) %>%
  mutate(n_ensembl_per_gene = length(ensembl_gene_id)) %>%
  ungroup() %>%
  # remove genes with ensembl id mapping to multiple or no genes
  filter(n_gene_per_ensembl == 1) %>%
  # take the sum of symbols that have multiple corresponding ensembl genes
  pivot_longer(cols = -c("ensembl_gene_id",
                         "external_gene_name",
                         "n_gene_per_ensembl",
                         "n_ensembl_per_gene"),
               names_to = "sample",
               values_to = "counts") %>%
  group_by(external_gene_name, sample) %>%
  mutate(counts_sum = sum(counts)) %>%
  dplyr::select(external_gene_name, sample, counts_sum) %>%
  distinct() %>%
  pivot_wider(id_cols = c("external_gene_name"),
              values_from = "counts_sum",
              names_from = "sample") %>%
  arrange(external_gene_name) %>%
  column_to_rownames("external_gene_name")


# ABUNDANCE
# filter ENSEMBL IDs for protein coding genes
ensembl_abundance_mouse_pc <- assay(salmon_se_mm, "abundance") %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  mutate(ensembl_gene_id = str_remove(ensembl_gene_id, "\\..*$")) %>%
  filter(ensembl_gene_id %in% protein_coding_genes_mouse$ensembl_gene_id) %>%
  column_to_rownames("ensembl_gene_id")

# filter and map to genenames
gene_abundance_mouse_pc <- assay(salmon_se_mm, "abundance") %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  mutate(ensembl_gene_id = str_remove(ensembl_gene_id, "\\..*$")) %>%
  # join (inner join includes filtering) and remove empty symbols
  inner_join(protein_coding_genes_mouse %>%
               filter(external_gene_name != ""),
             by = "ensembl_gene_id") %>%
  # handle ambiguously mapped genes
  group_by(ensembl_gene_id) %>%
  mutate(n_gene_per_ensembl = length(external_gene_name)) %>%
  group_by(external_gene_name) %>%
  mutate(n_ensembl_per_gene = length(ensembl_gene_id)) %>%
  ungroup() %>%
  # remove genes with ensembl id mapping to multiple or no genes
  filter(n_gene_per_ensembl == 1) %>%
  # take the sum of symbols that have multiple corresponding ensembl genes
  pivot_longer(cols = -c("ensembl_gene_id",
                         "external_gene_name",
                         "n_gene_per_ensembl",
                         "n_ensembl_per_gene"),
               names_to = "sample",
               values_to = "abundance") %>%
  group_by(external_gene_name, sample) %>%
  mutate(abundance_sum = sum(abundance)) %>%
  dplyr::select(external_gene_name, sample, abundance_sum) %>%
  distinct() %>%
  pivot_wider(id_cols = c("external_gene_name"),
              values_from = "abundance_sum",
              names_from = "sample") %>%
  arrange(external_gene_name) %>%
  column_to_rownames("external_gene_name")


rowdata_mm <- protein_coding_genes_mouse %>%
  filter(external_gene_name != "",
         external_gene_name %in% rownames(gene_counts_mouse_pc),
         ensembl_gene_id %in% str_remove(rownames(ensembl_counts_mouse_pc), "\\..*")) %>%
  group_by(external_gene_name) %>%
  summarise(mapped_ensembl_ids = paste(ensembl_gene_id, collapse = ", "),
            n_mapped = n()) %>%
  ungroup() %>%
  arrange(external_gene_name) %>%
  DataFrame()

coldata_hs <- colData(salmon_se_mm) 

stopifnot(all(rownames(gene_counts_mouse_pc) == rowdata_mm$external_gene_name))
stopifnot(all(rownames(gene_abundance_mouse_pc) == rowdata_mm$external_gene_name))

stopifnot(all(colnames(gene_counts_mouse_pc) == rownames(coldata_mm)))
stopifnot(all(colnames(gene_abundance_mouse_pc) == rownames(coldata_mm)))

salmon_se_mm_symbol <- SummarizedExperiment(
  assays = list(counts = gene_counts_mouse_pc,
                abundance = gene_abundance_mouse_pc),
  colData = coldata_mm,
  rowData = rowdata_mm
)
```

Here, samples from the correct tissues and conditions are selected for subsequent analyses.

```{r}
# filter to only include samples that will be used
# this includes: all inflammatory comp studies
# HC samples from either BM or PB from additional studies (non-inflammatory-comp)

salmon_se_hs_symbol_used <- salmon_se_hs_symbol[
  ,
  salmon_se_hs_symbol$inflammatory_analysis |
    (salmon_se_hs_symbol$origin_tissue %in% c("PB", "BM") & salmon_se_hs_symbol$condition == "HC")
]
salmon_se_mm_symbol_used <- salmon_se_mm_symbol[
  ,
  salmon_se_mm_symbol$inflammatory_analysis |
    (salmon_se_mm_symbol$origin_tissue %in% c("PB", "BM") & salmon_se_mm_symbol$condition == "HC")
]


# write coldata as meta table
coldata_hs_used <- salmon_se_hs_symbol_used %>%
  colData() %>%
  as_tibble()

coldata_mm_used <- salmon_se_mm_symbol_used %>%
  colData() %>%
  as_tibble()

write_tsv(
  coldata_hs_used,
  "data/metadata/table_S4_human.tsv"
)
write_tsv(
  coldata_mm_used,
  "data/metadata/table_S4_mouse.tsv"
)

# write summary files
coldata_hs_summary <- coldata_hs_used %>%
  group_by(study_id) %>%
  summarise(
    n = n(),
    inflammatory_analysis = unique(inflammatory_analysis),
    conds = paste(
      unique(
        paste0(origin_tissue, "_" ,condition)
        ),
      collapse = ", ")
    )
write_tsv(
  coldata_hs_summary,
  "data/metadata/summary_S1_human.tsv"
)

coldata_mm_summary <- coldata_mm_used %>%
  group_by(study_id) %>%
  summarise(
    n = n(),
    inflammatory_analysis = unique(inflammatory_analysis),
    conds = paste(
      unique(
        paste0(origin_tissue, "_" ,condition)
        ),
      collapse = ", ")
    )
write_tsv(
  coldata_mm_summary,
  "data/metadata/summary_S1_mouse.tsv"
)
```


# Save processed data for downstream use

```{r}
saveRDS(salmon_se_hs_symbol_used, "data/processed/arch_hs_se.rds")
saveRDS(salmon_se_mm_symbol_used, "data/processed/arch_mm_se.rds")
```
