---
title: "Inflammatory DE testing"
author: "NSH"
output: html_notebook
---

```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r, message=FALSE}
if (!require(EnhancedVolcano)) install.packages("EnhancedVolcano")
if (!require(gridExtra)) install.packages("gridExtra")

if (!require(BiocManager)) install.packages("BiocManager")
if (!require(DESeq2)) BiocManager::install("DESeq2")
if (!require(apeglm)) BiocManager::install("apeglm")

source("scripts/utils/utils.R")
source("scripts/utils/figure_builder.R")
```

# Inflammatory DE testing

This code performs differential expression testing for each selected study using DESeq2 and the lfcshrink algorithm.

## Sample selection and metadata representation

```{r}
cross_se <- readRDS("data/processed/cross_se.rds")
cross_se_inflammatory <- cross_se[, cross_se$inflammatory_analysis]
```

```{r}
study_stats <- cross_se_inflammatory %>%
  colData() %>%
  as.data.frame() %>%
  group_by(study_id) %>%
  summarise(
    species = unique(species),
    n_stimuli = length(unique(condition[condition != "HC"])),
    stimuli = paste(unique(condition[condition != "HC"]), collapse = ", "),
    n_samp_resting = sum(condition == "HC"),
    n_samp_stim = sum(condition != "HC"),
    n_tissues = length(unique(origin_tissue)),
    tissues = paste(unique(origin_tissue), collapse = ", "),
    model = unique(model)
    ) %>%
  filter(n_stimuli != 0)


saveRDS(study_stats, "data/processed/data_fig_4a.rds")
```


```{r}
fig_3a_1 <- build_fig_3a_1()
grid.draw(fig_3a_1)
```


### Experimental metadata representation

```{r}
fig_3a_2 <- build_fig_3a_2()
fig_3a_2
```


## Perform the testing

Having loaded the data, we can split our experiment into a list of experiments based on the study annotation.

```{r}
se_list <- split_se(cross_se_inflammatory, "series")[[1]]

# only keep datasets with more than one experimental condition
se_list_keep <- unlist(lapply(se_list, function (SEobject) return (length(unique(colData(SEobject)[["condition"]])) > 1)))

se_list_filtered <- se_list[se_list_keep]

for(i in 1:length(se_list_filtered)){
  print(paste0(colData(se_list_filtered[[i]])$species[1], length(unique(rownames(colData(se_list_filtered[[i]]))))))
}

perform_de_seq <- function (SEobject) {
  assays (SEobject) <- lapply(assays(SEobject), as.matrix)
  # dds <- DESeqDataSet(SEobject, design = ~ condition)
  dds <- DESeqDataSetFromMatrix(countData = round(assay(SEobject)),
                                colData = colData(SEobject),
                                rowData = rowData(SEobject),
                                design = ~ condition)
  # dds <- DESeqDataSetFromTximport(SEobject, design = ~ condition)
  
  if (!"HC" %in% levels(dds$condition)) {
    stop ("\n[DESEQ WRAPPER] HC not present in dataset and cannot be used as reference level.\n")
  }

  message ("\n[DESEQ WRAPPER] HC present, using HC as reference level.\n")
  dds$condition <- relevel(dds$condition, ref = "HC")
  # Pre-filtering the dataset
  nrow(dds)
  keep <- rowSums(counts(dds)) > 1
  dds <- dds[keep,]
  nrow(dds)
  
  dds <- DESeq(dds)
}

dds_list <- lapply (se_list_filtered, perform_de_seq)
```

We can now use our list of dds Objects to extract the results for our comparisons of interest.

```{r}
extract_results <- function (dds) {
  message ("[RES EXTRACT] Levels present in dataset: ", paste(levels(dds$condition), collapse = ", "))

  if (length(levels(dds$condition)) > 2) {
    message ("[RES EXTRACT] More than two pathological conditions found.")
  }
  cond_list <- as.list (setdiff(levels(dds$condition), "HC"))
  # TODO: We might also want to add a unique study suffix such as GSE
  species_list <- as.list(unique(dds$species))
  # TODO: check naming (space vs _)
  study_str <- unique (dds$study_id)
  
  if (length(study_str) != 1) {
    warning ("[RES EXTRACT] More than one study present in dataset: ", paste(study_str, collapse = ", "))
    study_str <- paste(study_str, collapse = "_")
  }
  
  res_list <- lapply (cond_list, function (condition, dds, study_str) {
    coef_str <- paste0("condition_", condition, "_vs_HC")
    # TODO: coef vs. contrast
    warning(print(study_str))
    res <- lfcShrink(dds, coef = coef_str, type = "apeglm")
  }, dds, study_str)
  
  names (res_list) <- paste0(study_str, "_condition_", as.character(cond_list), "_vs_HC_", species_list)
  
  return (res_list)
  
}

res_list <- lapply (dds_list, extract_results)
# flatten list
res_list <- unlist (res_list)
```

## Plot the resulting volcanos

```{r, fig.width=20, fig.height=25}
plts <- lapply (seq_along(res_list), function(i, res_list) {
  
  ttable <- res_list[[i]]
  
  plt <- EnhancedVolcano(ttable,
                         title = names(res_list)[[i]],
                         lab = rownames(ttable),
                         x = "log2FoldChange",
                         y = "padj",
                         pCutoff = 0.05,
                         FCcutoff = 1)
  }, res_list)
#grid.arrange(grobs = plts)
```

```{r, fig.height=5, fig.width=3}
lapply(plts, function(plt){plt})
```


Save the results.

```{r}
saveRDS(res_list, "data/processed/de_seq_res_list.rds")
```
