---
title: "Regulator enrichment for inflammatory fisher signature"
author: "NSH, FAR"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r, message=FALSE, include=FALSE}
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(magrittr)) install.packages("magrittr")
if (!require(dorothea)) BiocManager::install("dorothea")
if (!require(decoupleR)) BiocManager::install("decoupleR")
if (!require(OmnipathR)) BiocManager::install("OmnipathR")


source("scripts/utils/config.R")
source("scripts/utils/utils.R")
source("scripts/utils/figure_builder.R")
```

```{r, include=FALSE}
res_list <- readRDS("data/processed/de_seq_res_list.rds")
lfc_df <- readRDS("data/processed/fisher_lfc_df.rds")
```


# Identify regulators that are associated with gene upregulation

```{r}
upreg_genes_list <- lapply (res_list, function (res) {
  
  upreg_genes <- res %>%
    as.data.frame() %>%
    rownames_to_column("symbol") %>%
    filter(log2FoldChange > 0.5, padj < 0.05) %>% 
    slice_min(order_by = padj, n = 250, with_ties = F) %>%
    pull(symbol)
  
})

chea_res_up <- lapply (names(upreg_genes_list), function (cur_name) {
  
  message("Chea...")
  get_chea(upreg_genes_list[[cur_name]], cur_name)
  
})

chea_res_up_df <- lapply (chea_res_up, function (res) {
  
  return (res[[1]])
  
})

chea_res_up_df_all <- Reduce(rbind, chea_res_up_df)

saveRDS(chea_res_up_df_all, "data/processed/data_fig_5a_1.rds")
```

Explore results on an individual level.

```{r}
nqueries <- length(unique(chea_res_up_df_all$`Query Name`))
chea_matrix <- chea_res_up_df_all %>%
  transmute(qname = `Query Name`,
            TF,
            logscore = -log10(as.numeric(Score))) %>%
  group_by(TF) %>%
  mutate(mean_tf_activity = mean (logscore)) %>%
  ungroup() %>%
  slice_max(order_by = mean_tf_activity, n = 40*nqueries, with_ties = FALSE) %>%
  dplyr::select(-c(mean_tf_activity)) %>%
  pivot_wider(names_from = "qname",
              values_from = "logscore") %>%
  column_to_rownames("TF")

chea_matrix <- chea_matrix %>%
  t() %>%
  scale() %>%
  t()
  
Heatmap(chea_matrix,
        cluster_rows = FALSE,
        cluster_columns = FALSE)
```

Look at correlation of individual TF activity.

```{r, fig.height=5, fig.width=5}
explore_up <- chea_res_up_df_all %>%
  transmute(qname = `Query Name`,
            TF,
            logscore = -log10(as.numeric(Score))) %>%
  pivot_wider(names_from = "qname",
              values_from = "logscore") %>%
  column_to_rownames("TF")

expl_up_cor <- cor(explore_up, explore_up)
Heatmap(expl_up_cor)
```

Scatter plot species comparison.

```{r, fig.width=5, fig.height=5}
fig_4b <- build_fig_4b()

fig_4b 
```

# Identify regulators that are associated with gene downregulation

```{r}
dnreg_genes_list <- lapply (res_list, function (res) {
  
  dnreg_genes <- res %>%
    as.data.frame() %>%
    rownames_to_column("symbol") %>%
    filter(log2FoldChange < -0.5, padj < 0.05) %>% 
    slice_min(order_by = padj, n = 250, with_ties = F) %>%
    pull(symbol)
  
})

chea_res_dn <- lapply (names(dnreg_genes_list), function (cur_name) {
  
  message("Chea...")
  get_chea(dnreg_genes_list[[cur_name]], cur_name)
  
})

chea_res_dn_df <- lapply (chea_res_dn, function (res) {
  
  return (res[[1]])
  
})

chea_res_dn_df_all <- Reduce(rbind, chea_res_dn_df)
saveRDS(chea_res_dn_df_all, "data/processed/data_fig_5a_2.rds")
```

Explore results on an individual level.

```{r, fig.height=5, fig.width=5}
explore_dn <- chea_res_up_df_all %>%
  transmute(qname = `Query Name`,
            TF,
            logscore = -log10(as.numeric(Score))) %>%
  pivot_wider(names_from = "qname",
              values_from = "logscore") %>%
  column_to_rownames("TF")

expl_cor_dn <- cor(explore_dn, explore_dn)
Heatmap(expl_cor_dn)
```

Scatter plot species comparison.

```{r, fig.width=4, fig.height=4}
fig_4a <- build_fig_4a()

fig_4a
```


# Regulatory activity profiles

Plot the chea scores for each individual comparisons most up- and down-
regulated genes.
To this end we first perform two chea queries per comparison (up and down
respectively) which we then combine for plotting.

## Get ChEA data

```{r}
chea_res_updn_df_all <- rbind(
  chea_res_up_df_all %>%
    mutate(`Query Name` = paste(`Query Name`, "_up")),
  chea_res_dn_df_all %>%
    mutate(`Query Name` = paste(`Query Name`, "_dn"))
)
```

## add some statistical reasoning for TF display

```{r, fig.height=10, fig.width=10}
TF_stats <- chea_res_updn_df_all %>%
  transmute(qname = `Query Name`,
            TF,
            score = as.numeric(Score),
            group = str_remove(qname, ".*_"),
            comp = str_remove_all(qname, "_up|_dn")) %>%
  dplyr::select(-qname) %>%
  pivot_wider(names_from = group,
              values_from = score,
              names_prefix = "score_") %>%
  mutate(diff = score_up - score_dn)

ggplot(TF_stats, aes(x = diff)) +
  geom_histogram()

example_tfs <- TF_stats %>%
  pull(TF) %>%
  unique() %>%
  `[`(1:100)

ggplot(TF_stats %>%
         filter(TF %in% example_tfs),
       aes(x = diff)) +
  geom_density() +
  facet_wrap(~TF)
```

```{r}
TF_stats %<>%
  group_by(TF) %>%
  mutate(p = t.test(score_up,
                    score_dn,
                    alternative = "two.sided",
                    paired = T)$p.value,
         t = t.test(score_up,
                    score_dn,
                    alternative = "two.sided",
                    paired = T)$statistic,
         meandiff = t.test(score_up,
                           score_dn,
                           alternative = "two.sided",
                           paired = T)$estimate) %>%
  ungroup()

TF_stats_summary <- TF_stats %>%
  group_by(TF) %>%
  summarise(TF = unique(TF),
            p = unique(p),
            t = unique(t),
            meandiff = unique(meandiff)
            ) %>%
  mutate(padj = p.adjust(p, method = "BH")) %>%
  arrange(padj)

TF_stat_row_annotation <- TF_stats_summary %>%
  filter(padj < 0.05)  %>%
  transmute(TF,
            tf_group = case_when(
              t < 0 ~ "a_up",
              t > 0 ~ "c_dn"
              ),
            group_score = t,
            padj
            ) %>%
  arrange(group_score) %>%
  column_to_rownames("TF")
```



## Prepare data for plotting

```{r}
plt_data <- chea_res_updn_df_all %>%
  transmute(qname = `Query Name`,
            TF,
            logscore = -log10(as.numeric(Score))) %>%
  pivot_wider(names_from = "qname",
              values_from = "logscore") %>%
  column_to_rownames("TF")

col_annotation <- plt_data %>%
  rownames_to_column("TF") %>%
  pivot_longer(cols = -"TF",
               names_to = "qname",
               values_to = "logscore") %>%
  mutate(set = str_remove(qname, "^.*_")) %>%
  dplyr::select(qname, set) %>%
  distinct() %>%
  arrange(set) %>%
  column_to_rownames("qname")

row_annotation <- TF_stat_row_annotation


# filter and arrange
plt_data <- plt_data[rownames(plt_data) %in% rownames(row_annotation), ]
plt_data <- plt_data[match(rownames(row_annotation), rownames(plt_data)),
                         match(rownames(col_annotation), colnames(plt_data))]

saveRDS(plt_data, "data/processed/data_1_fig_5c.rds")
saveRDS(col_annotation, "data/processed/data_2_fig_5c.rds")
saveRDS(row_annotation, "data/processed/data_3_fig_5c.rds")
```


## Plot activity heatmap

```{r fig.height=3, fig.width=5}
fig_4d <- build_fig_4d()

fig_4d
```

Dorothea analysis: 
(1) Plot logFC of TF expression, use individual databases per species (dorothea) to avoid omitting anything.

```{r}
lfc_means <- lfc_df %>%
  mutate(mean_Mm = rowMeans(.[,str_sub(colnames(lfc_df), start = -2) == "Mm" ]),
         mean_Hs = rowMeans(.[,str_sub(colnames(lfc_df), start = -2) == "Hs" ])) %>%
  dplyr::select(symbol, mean_Hs, mean_Mm)

ortho_human_to_mouse <- readRDS("data/processed/ortho_human_to_mouse.rds")
ortho_mouse_to_human <- readRDS("data/processed/ortho_mouse_to_human.rds")

lfc_means <- merge(lfc_means, ortho_human_to_mouse[,c("external_gene_name", "mmusculus_homolog_associated_gene_name")],
                   by.x = "symbol", by.y = "external_gene_name", all.x = T)

lfc_means_tf <- lfc_means %>%
  filter(symbol %in% dorothea_hs$tf | mmusculus_homolog_associated_gene_name %in% dorothea_mm$tf)

saveRDS(lfc_means_tf, file = "data/processed/data_fig_4e.rds")
```


(2) Perform TFEA using decoupleR (based on dorothea databases) per species, then merge and prepare for plots.

```{r}
lfc_df %>%
  mutate(mean_Mm = rowMeans(.[,str_sub(colnames(lfc_df), start = -2) == "Mm" ]),
         mean_Hs = rowMeans(.[,str_sub(colnames(lfc_df), start = -2) == "Hs" ])) %>%
  dplyr::select(symbol, mean_Hs, mean_Mm)

lfc_df_Hs <- lfc_df %>%
  column_to_rownames("symbol")
lfc_df_Hs <- as.matrix(lfc_df_Hs[,str_sub(colnames(lfc_df_Hs), start = -2) == "Hs" ])

lfc_df_Mm <- lfc_df %>%
  left_join(ortho_human_to_mouse[,c("external_gene_name", "mmusculus_homolog_associated_gene_name")],
            by = c("symbol" = "external_gene_name")) %>%
  dplyr::select(-symbol) %>%
  column_to_rownames("mmusculus_homolog_associated_gene_name")
lfc_df_Mm <- as.matrix(lfc_df_Mm[,str_sub(colnames(lfc_df_Mm), start = -2) == "Mm" ])


tf_Hs <- get_dorothea(organism='human', levels=c('A', 'B', 'C'))
tf_Mm <- get_dorothea(organism='mouse', levels=c('A', 'B', 'C'))


act_Hs <- run_wmean(mat=lfc_df_Hs, net=tf_Hs, .source='source', .target='target',
                    .mor='mor', times = 100, minsize = 5)

act_Mm <- run_wmean(mat=lfc_df_Mm, net=tf_Mm, .source='source', .target='target',
          .mor='mor', times = 100, minsize = 5)

act_merged_Hs <- act_Hs %>%
  left_join(ortho_human_to_mouse[,c("external_gene_name", "mmusculus_homolog_associated_gene_name")], by = c("source" = "external_gene_name")) %>%
  mutate(symbol_hs = source) %>%
  dplyr::rename(symbol_mm = mmusculus_homolog_associated_gene_name) %>%
  dplyr::select(symbol_hs, symbol_mm, condition, score, p_value) %>%
  mutate(species = "Hs")

act_merged_Mm <- act_Mm %>%
  left_join(ortho_mouse_to_human[,c("external_gene_name", "hsapiens_homolog_associated_gene_name")], by = c("source" = "external_gene_name")) %>%
  mutate(symbol_mm = source) %>%
  dplyr::rename(symbol_hs = hsapiens_homolog_associated_gene_name) %>%
  dplyr::select(symbol_hs, symbol_mm, condition, score, p_value) %>%
  mutate(species = "Mm")

act_merged <- rbind(act_merged_Hs, act_merged_Mm)

saveRDS(act_merged, file = "data/processed/data_fig_4f.rds")
```

